Evaluation of pilot jobs for Apache Spark\\ applications on HPC clusters

### 2019-05-16 paper.tex

\begin{document}
\title{Evaluation of pilot jobs for Apache Spark\\ applications on HPC clusters}
\author{
    \IEEEauthorblockN{
        Val\'erie Hayot-Sasson and Tristan Glatard \\
        Department of Computer Science and Software Engineering, Concordia University, Montreal, Canada
    }
    \IEEEauthorblockA{}
}
\maketitle

\begin{abstract}
    Big Data has become prominent throughout many scientific fields and, as
    a result, scientific communities have sought out Big Data frameworks to
    accelerate the processing of their increasingly data-intensive
    pipelines. However, while scientific communities typically rely on
    High-Performance Computing (HPC) clusters for the parallelization of
    their pipelines, many popular Big Data frameworks such as Hadoop and
    Apache Spark were primarily designed to be executed on dedicated
    commodity infrastructures. This paper evaluates the benefits of pilot
    jobs over traditional batch submission on HPC clusters with overlay
    Apache Spark clusters. We first evaluate the overall speedup brought on
    by employing pilot jobs. We then examine the robustness to master and
    driver failures, which may be frequent when underestimating pilot
    walltime. Spark built-in checkpointing is also investigated in relation
    to pilot expiration. Surprisingly, our results show that the speed-up
    provided by pilot jobs over batch scheduling remains moderate (\TG{X} on
    average) despite the presence of long queuing times. In addition, pilot
    jobs provide an extra layer of scheduling that complexifies debugging
    and deployment. We conclude that traditional batch scheduling should
    remain the default strategies to deploy Apache Spark applications on
    HPC clusters.
\end{abstract}

@@

        \todo{scala standalone vs our pyspark workaround. kill masters in experiments}
    \subsection{Checkpointing}
        \todo{metric for determining how often to checkpoint based on cluster size}
    \subsection{Job arrays}
        \todo{need to kill idle workers. may not want all workers to be running at once.}
    \subsection{Example application}
        \todo{incrementation with varying task durations}
\section{Results}\label{sec:results}
\section{Discussion}\label{sec:discussion}

SPA vs other frameworks.

Technical overhead of using pilots + an overlay cluster over a cluster.

>>>>

\begin{document}

\title{Evaluation of pilot jobs for Apache Spark\\ applications on HPC clusters}
\author{
    \IEEEauthorblockN{
        Val\'erie Hayot-Sasson and Tristan Glatard \\
        Department of Computer Science and Software Engineering, Concordia University, Montreal, Canada
    }
    \IEEEauthorblockA{}
}
\maketitle

\begin{abstract}
    Big Data has become prominent throughout many scientific fields and, as
    a result, scientific communities have sought out Big Data frameworks to
    accelerate the processing of their increasingly data-intensive
    pipelines. However, while scientific communities typically rely on
    High-Performance Computing (HPC) clusters for the parallelization of
    their pipelines, many popular Big Data frameworks such as Hadoop and
    Apache Spark were primarily designed to be executed on dedicated
    commodity infrastructures. This paper evaluates the benefits of pilot
    jobs over traditional batch submission on HPC clusters with overlay
    Apache Spark clusters. We first evaluate the overall speedup brought on
    by employing pilot jobs. We then examine the robustness to master and
    driver failures, which may be frequent when underestimating pilot
    walltime. Spark built-in checkpointing is also investigated in relation
    to pilot expiration. Surprisingly, our results show that the speed-up
    provided by pilot jobs over batch scheduling remains moderate (\TG{X} on
    average) despite the presence of long queuing times. In addition, pilot
    jobs provide an extra layer of scheduling that complexifies debugging
    and deployment. We conclude that traditional batch scheduling should
    remain the default strategies to deploy Apache Spark applications on
    HPC clusters.

\end{abstract}

@@

        \todo{scala standalone vs our pyspark workaround. kill masters in experiments}
    % maybe remove\subsection{Checkpointing}
        \todo{metric for determining how often to checkpoint based on cluster size}
    % maybe remove \subsection{Job arrays}
        \todo{need to kill idle workers. may not want all workers to be running at once.}
    % maybe remove \subsection{Example application}
        \todo{incrementation with varying task durations}

\section{Results}\label{sec:results}

As can be seen in both figures~\ref{fig:makespansbeluga} and ~\ref{fig:makespanscedar}, pilots 
generally did not bring any performance improvement to the execution of Spark workloads
on Beluga and Cedar. In Beluga (Figure~\ref{fig:beluga1}), 16 pilots were always 
slower than batch except for repetitions 4 and 12. The maximum speedup achieved
was approximately 1.3~x for 8 pilots and 1.1~x for 16 pilots. In Figures~\ref{fig:beluga2}
and ~\ref{fig:beluga3}, it is also the case that batch is generally faster than
pilots for the majority of the cases, except a couple, in which the speedup brought
by use of pilots is very limited. It is only in Configuration~4 (Figure~\ref{fig:beluga4})
that we see a maximum speedup of 2.8, however, it generally remains that pilots are slower.

The Cedar cluster (Figure~\ref{fig:makespanscedar}) also follows the same trend that
the makespan of batch scheduled applications are generally faster than that of pilots,
with a maximum speedup of 3, found both in Configuration~1 and 4. Moreover, in 
Cedar, it was less clear whether to use 8 or 16 pilots as speedup between the two
varies between the iterations. In general, both pilot configurations performed the
same.

In order to determine whether queueing time was the cause of the difference 
in makespans between batch and pilots, we compared makespan with queuing time of 
first pilot. As can be seen in Figure~\ref{fig:mqcedar}, variations in overall
makespans is primiarily due to queueing time.

In the case of pilots, multiple Slurm jobs, each with potentially different queuing times,
are used. Figure~\ref{fig:mwall} displays the makespan times for an average number of workers
given an experiment configuration, where the line corresponds to the estimated makespan using 
our model. Makespan for the model was calculated as:
$$
M = (\lceil 125/W \rceil\times d) + 20,
$$
where:
\begin{itemize}
    \item $M$ is the makespan of the application
    \item $W$ is the average number of Spark workers throughout the execution
    \item $d$ is the task delay associated with a given configuration
    \item $20$ is the average measured incrementation duration for a BigBrain chunk
\end{itemize}
As can be seen in the figure, although pilots and batch had similar makespans
for a given number of average workers, pilots were consistently slower than batch
for the same number of workers. Both batch and pilots were consistent with the 
model, however. In general, both pilots and batch had the same number of average
workers.

    \begin{figure*}
        \centering
        \begin{subfigure}[b]{0.475\textwidth}
            \centering
            \includegraphics[width=\textwidth]{figures/dedicated_1_beluga}
            \caption[]%
            {{\small Configuration 1}}
            \label{fig:beluga1}
        \end{subfigure}
        \hfill
        \begin{subfigure}[b]{0.475\textwidth}
            \centering
            \includegraphics[width=\textwidth]{figures/dedicated_2_beluga}
            \caption[]%
            {{\small Configuration 2}}
            \label{fig:beluga2}
        \end{subfigure}
        \vskip\baselineskip
        \begin{subfigure}[b]{0.475\textwidth}
            \centering
            \includegraphics[width=\textwidth]{figures/dedicated_3_beluga}
            \caption[]%
            {{\small Configuration 3}}
            \label{fig:beluga3}
        \end{subfigure}
        \quad
        \begin{subfigure}[b]{0.475\textwidth}
            \centering
            \includegraphics[width=\textwidth]{figures/dedicated_4_beluga}
            \caption[]%
            {{\small Configuration 4}}
            \label{fig:beluga4}
        \end{subfigure}
        \caption[]
        {\small The application makespan on the Beluga cluster for all repetitions.}
        \label{fig:makespansbeluga}
    \end{figure*}

    \begin{figure*}
        \centering
        \begin{subfigure}[b]{0.475\textwidth}
            \centering
            \includegraphics[width=\textwidth]{figures/dedicated_1_cedar}
            \caption[]%
            {{\small Configuration 1}}
            \label{fig:cedar1}
        \end{subfigure}
        \hfill
        \begin{subfigure}[b]{0.475\textwidth}
            \centering
            \includegraphics[width=\textwidth]{figures/dedicated_2_cedar}
            \caption[]%
            {{\small Configuration 2}}
            \label{fig:cedar2}
        \end{subfigure}
        \vskip\baselineskip
        \begin{subfigure}[b]{0.475\textwidth}
            \centering
            \includegraphics[width=\textwidth]{figures/dedicated_3_cedar}
            \caption[]%
            {{\small Configuration 3}}
            \label{fig:cedar3}
        \end{subfigure}
        \quad
        \begin{subfigure}[b]{0.475\textwidth}
            \centering
            \includegraphics[width=\textwidth]{figures/dedicated_4_cedar}
            \caption[]%
            {{\small Configuration 4}}
            \label{fig:cedar4}
        \end{subfigure}
        \caption[]
        {\small The application makespan on the Cedar cluster for all repetitions.}
        \label{fig:makespanscedar}
    \end{figure*}

      \begin{figure*}
        \centering
        \begin{subfigure}[b]{0.475\textwidth}
            \centering
            \includegraphics[width=\textwidth]{figures/dedicated_4_cedar}
            \caption[]%
            {{\small Makespan}}
            \label{fig:makespancedar}
        \end{subfigure}
        \hfill
        \begin{subfigure}[b]{0.475\textwidth}
            \centering
            \includegraphics[width=\textwidth]{figures/pq_4_cedar}
            \caption[]%
            {{\small Queuing time}}
            \label{fig:queuecedar}
        \end{subfigure}
        \caption[]
        {\small Side-by-side comparison of total makespan and queueuing time of
        first pilots or batch application for Configuration 4 running on Cedar}
        \label{fig:mqcedar}
    \end{figure*}

    \begin{figure*}
        \centering
        \begin{subfigure}[b]{0.475\textwidth}
            \centering
            \includegraphics[width=\textwidth]{figures/mw_beluga}
            \caption[]%
            {{\small Beluga}}
            \label{fig:mwbeluga}
        \end{subfigure}
        \hfill
        \begin{subfigure}[b]{0.475\textwidth}
            \centering
            \includegraphics[width=\textwidth]{figures/mw_cedar}
            \caption[]%
            {{\small Cedar}}
            \label{fig:mwcedar}
        \end{subfigure}
        \caption[]
        {\small The relation between makespan and average number of workers as 
        calculated using Equation~\ref{eq:avgw}. Trendline denotes the expected makespan
        given the average number of workers given a certain configuration}
        \label{fig:mwall}
    \end{figure*}

\section{Discussion}\label{sec:discussion}

%SPA vs other frameworks.

%Technical overhead of using pilots + an overlay cluster over a cluster.
Pilots do not appear to provide any performance advantage over batch scheduling.
This can be due to a few reasons. The queuing time for batch and pilots were similar, 
which can be an indication that the required resources to run the batch requests were
available, and our priority on the cluster was what affected the queuing time. This may, in turn,
explain why batch was sometimes significantly slower than pilots -- queueing time was not
only controlled by priority, but also by available resources.

It was also found that having 16 pilots was generally slower than 8 for the same
configuration. This was particularly the case on Beluga. We suspect this might be
related to the number of Slurm jobs permitted for a user on the backfill queue. For
all configurations, we ran batch, 8 pilots and 16 pilots concurrently. This would have
resulted in 25 Slurm jobs trying the access the queue concurrently. On Beluga, this
feature is configured to a total of 10 jobs. For 16 nodes, this would at best be
only 62.5\% of the pilots being scheduled at a given time if all resources are available.
The next 6 jobs would then have to wait 3 minutes before having an opportunity to enter the
backfill. Cedar, on the other hand, permits 40 user jobs to be backfilled at a given time.
Therefore, on Cedar, all concurrent Slurm jobs had an opportunity to be backfilled at
the same time, which may explain why this behaviour is more apparent on B\'eluga. Furthermore,
while both pilot configurations could be places in B\'eluga's low memory nodes, batch requests
had to be place on B\'eluga's medium memory nodes. B\'eluga has 172 low memory nodes and
516 mid memory nodes. As pilots would get priority for the low memory nodes first and 
low memory nodes are less frequent that medium memory nodes, it is possible that this
may have increased the overall queuing times of the pilots. However, Cedar's basic
node has enough memory to for all batch and pilot configuration, therefore, this could not
have affected queuing time for Cedar.

Although queuing time is largely responsible for makespan variations, it is not
always entirely responsible for the difference. Sometimes there are errors related
to workers not registering properly with the master. These occurences lead to the 
pipeline being processed entirely with a smaller number of available workers. As the task
delay added considers the maximum amount of parallel workers, functioning with less total workers
will significantly affect makespan. However, as seen in Figure~\ref{fig:mwall}
pilots are slower even with the same amount of average workers. A potential reasoning for this
may be the, worker registration delay and the time required to transfer data
over after the application has already started. In batch scheduling, all workers a started in
parallel and the data is transferred automatically at the beginning of the pipeline as the maximum amount
of parallelism is available at driver start time. Pilots, on the other hand, may start workers in parallel,
but not necessarily all will be started at the same time. This means that the entire 
workflow may suffer the impacts of starting workers in sequence. Moreover, there would also
be a delay with respect to transferring data from workers to newly added workers. Therefore,
it is only natural that pilots would have a bit more overhead that batch given the same
queuing time should pilots not all commence at once. Furthermore, the pilot scripts also
have some starter overhead as the all start a master, a worker (registered to the main
master) and attempt to start a driver. Batch need only start a single master and attempts to start
a driver only once.

The motivation to use pilots, however, is dynamic scheduling on HPC clusters. While we
investigated the queuing time differences with respect to differences in available resources,
the walltimes were kept static. Had walltime been underestimated to the point where the Slurm jobs
would terminate prior to application completion, pilots would have been preferable to batch
as batch scheduling has no mechanism to restart. Even if such a mechanism was in place, the entire
cluster would be shutdown and need to restart, which has additional overheads. With pilots, as long as 
there was at least one pilot alive, it would be possible to maintain the cluster and add additional pilots.
Spark provides built-it fault-tolerance for not only the workers, but the master and driver as well.
